---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

In activity data gathered from fitness sensors, we discover a previously unimagined 
pattern ofactivity with disturbing implication for mental health practitioners: 
Early morning exercise.

## Loading and preprocessing the data

```{r, echo=TRUE}
data = read.csv( "activity.csv" )
dim(data)
summary( data )

s1 <- sum( data$steps, na.rm = TRUE )
```

## What is mean total number of steps taken per day?

```{r}
totalStepsByDate <- aggregate( steps ~ date, data, sum )$steps
s2 <- sum( totalStepsByDate, na.rm = TRUE )
if( s1 != s2 ) stop( "Checksum mismatch");
hist( totalStepsByDate, bins <- 15) 
mean( totalStepsByDate )
median( totalStepsByDate )

```

This histogram reveals a roughly Gaussian distribution of daily activity levels, 
consistent with a model that views the daily activity level as a sum of 
multiple independent variables.

## What is the average daily activity pattern?

```{r}

aggregateDataByInterval <- function( df ) {
  aggregate( steps ~ interval, df, mean)
}

byIntervalPlot <- function( aggregatedDf) {
  plot( c(1:length( aggregatedDf$steps)) / 12, aggregatedDf$steps, type="l", col="blue", 
        ylim = c(0,250), ylab = "Mean steps", 
        xlab="Hour of day")
}

meanStepsByInterval <- aggregateDataByInterval( data )
summary( meanStepsByInterval )
byIntervalPlot( meanStepsByInterval )
title( main="Mean step count by 5 minute interval")


#barplot( meanStepsByInterval$steps )

#maximizer and maximum
meanStepsByInterval[ which.max( meanStepsByInterval$steps ),]

# A table of intervals with relatively high mean step counts
meanStepsByInterval[ meanStepsByInterval$steps > 90, ]

```

This plot shows high activity during the 8 a.m. hour, a distinct peak of activity 
around 8:35 a.m.. Otherwise, we note a moderate level 
of activity between 6 a.m. and 7 p.m with a few local maxima, and 
almost no activity bewteen 11 p.m. and 5 a.m.. 

These data do not enable us to assess whether high morning level is 
due to a large number of neurotic morning exercisers, whether it
reflects the pathologically elevated activity levels of a relatively small number 
of such troubled individuals, or a combination of the two. Resolving this 
key question appears problematic, owing to the difficulty of conducting 
reliable research in such grueling morning conditions.

## Imputing missing values

The missing data in the dataset consists of entire days for which 
no data was reported on any 5-minute interval, as the calculations below demonstrate.
We discern no simple pattern in the missing days.

```{r}
missingData <- data[ is.na( data$steps ), ]

missingByDate <- table( missingData$date )

which( missingByDate > 0 )
which( missingByDate >0 && missingByDate != 288 )

```

Accordintly, we model missing step counts as the corresponding average 
for the same interval for days with data. 

```{r}

# Compute the index ( 1:288 ) of a given interval.
# interval "210" is the 3rd inteval of the second hour: 0-55, 100-155, 200, 205, 210"
# So intervalNum( 210 ) -> 27.

intervalNum <- function( interval ) {
  hour <- floor( interval / 100 )
  min <- interval - hour * 100
  hour * 12 + min / 5 + 1
}

# For missing data, default to the mean for the corresponding interval
imputedValue <- function( steps, interval ) { 
  ifelse( is.na(steps), meanStepsByInterval[intervalNum(interval), 2], steps ) 
}

# Add column for data merged with imputed data.
data$imputedSteps <- mapply( imputedValue, data$steps, data$interval )

# Test: missing values are imputed, since first day has no data
norm( as.matrix( data$imputedSteps[1:288] - meanStepsByInterval[1:288,2] ) ) # -> 0
# Test: present values are left alone
norm( as.matrix( data$imputedSteps[2400:2430] - data$steps[2400:2430] ) ) # -> 0

imputedTotalStepsByDate <- aggregate( imputedSteps ~ date, data, sum )$imputedSteps

hist( imputedTotalStepsByDate, bins <- 15)

```

## Are there differences in activity patterns between weekdays and weekends?

To answer this question, we introduce a factor that is 1 on weekdays and 2 on weekends.

```{r, fig.height=8}
isWeekday <- function( dateString ) {
  weekdays( as.POSIXlt( dateString ) ) %in% c( "Saturday", "Sunday" )
}

# Create weekend/weekday factor, split on it, and label the splits frames.
data$isWeekday <- mapply( isWeekday, data$date)
splitData <- split( data, data$isWeekday )
weekdayData <- splitData[[1]]
weekendData <- splitData[[2]]

# Difference in means for split
weekdayMean <- mean( weekdayData$steps, na.rm = TRUE )
weekendMean <- mean( weekendData$steps, na.rm = TRUE )
weekendLevelIncrease <- round( 100 * (weekendMean - weekdayMean ) / weekdayMean, 2 )

# Compute average steps per interval for split
meanStepsByIntervalWeekday <- aggregateDataByInterval( weekdayData )
meanStepsByIntervalWeekend <- aggregateDataByInterval( weekendData )

# Panel plot
par( mfrow=c(2,1))
byIntervalPlot( meanStepsByIntervalWeekday )
title( main = "Weekdays: Mean step count by 5 minute interval" )
byIntervalPlot( meanStepsByIntervalWeekend )
title( main = "Weekends: Mean step count by 5 minute interval" )

```

In summary, on Saturdays and Sundays, the average activity level 
increases `r weekendLevelIncrease` from `r round(weekdayMean,2)` to `r round(weekendMean,2)` 
steps per interval. On weekends, the pathological early morning activity, 
while still significant, is reduced, the activity apparently distributed across later 
parts of the day and into the evening.
